Potential Factors of Decreasing Birth Rates in the US since the Great Recession and Beyond

### 2024-02-15 outputs/paper/US_birth_rates.qmd
Rename outputs/US_birth_rates.qmd to outputs/paper/US_birth_rates.qmd

---
title: "Potential Factors of Decreasing Birth Rates in the US since the Great Recession and Beyond"
subtitle: "Reproduction of 'The Puzzle of Falling US Birth Rates since the Great Recession'(Kearney, Levine & Pardue, 2022)"
author: 
  - Chay Park
thanks: "Code and data are available at: https://github.com/Chay-HyunminPark/Social-Science-Study. A replication of various aspects in this paper are available at: https://doi.org/10.48152/ssrp-srs6-t802""
date: today
date-format: long
abstract: "This paper examines the notable decrease in US birth rates from 2007 to 2020, replicating the study conducted by Kearney, Levine, and Pardue (2022). Through an analysis of trends spanning from 1980 to 2020, we investigate variations in birth rates among different age groups and disparities based on the race and ethnicity of mothers. While initially attributing the decline to the Great Recession, our analysis uncovers additional factors contributing to this trend. Furthermore, our exploration delves into socio-cultural aspects, offering insights for policymakers and researchers interested in population trends and their implications."
format: pdf
number-sections: true
bibliography: references.bib
prefer-html: true
---

@@

According to the Centers for Disease Control and Prevention, the birth rate is defined as "the number of live births per 1,000 population." This rate is calculated by dividing the number of live births in a population in a year by the midyear resident population. Specifically, for census years, rates are calculated based on the unrounded census counts of the resident population as of April 1st. From an economic perspective, birth rates are an important factor that determines the growth of a country. Recent years have seen widespread concern over declining birth rates globally, prompting nations to explore strategies to address this issue. For instance, Germany's birth rate in 2024 is 9.3, which is lower than Canada's 10.0 and the US's 12.0 (Macrotrends 2024). Germany is taking action by actively recruiting skilled workers from abroad (Grunau 2023) and reforming its citizenship law to allow dual citizenship (Knight 2024) as some measures to combat declining birth rates and overall population. Furthermore, birth rates serve as an indicator of replacement-level fertility, essential for maintaining population stability in developed nations (Craig 1994).

@@

This paper utilized a replication package of the original study and employed a cleaned data CSV file to reproduce the related figures and tables. While the original code in the replication package was written in Stata, this paper utilized R (R Core Team 2020) for analysis, leveraging packages such as ggplot2 (Wickham et al. 2016) for creating graphical representations. Additionally, the gridExtra package was used to enhance the layout and presentation of the figures (Murrell 2021).

The paper proceeds by first presenting the trend in US birth rates over the period 1980 to 2020. It then delves into variations in birth rates by age group and race and ethnicity of mothers, correlating these trends with education, socio-economic factors, and financial stability. The replication of Kearney, Levine, and Pardue's study focuses on answering the following research questions:

@@

Regarding race and ethnicity, "white" refers to the non-Hispanic Caucasian population group, "black" refers to the non-Hispanic African American group, and "Hispanic" refers to people with Hispanic ethnicity, mainly from South America, for brevity.

@@

# Read the data from CSV file
data <- read.csv(file = here::here("outputs/data/fig_1.csv"))

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-polscatter-graph
#| fig-cap: "Trend in US Birth Rates during the presidential mandate"

@@

# Read the data from CSV file
data <- read.csv(file = here::here("outputs/data/figs_2a_2b.csv"))

# Plotting the first line graph (Graph A)
graph_A <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = brate_1519, color = "blue"), size = 0.4) +
  geom_line(aes(y = brate_2024, color = "orange"), size = 0.4) +
  geom_line(aes(y = brate_2529, color = "grey"), size = 0.4) +
  geom_line(aes(y = brate_3034, color = "yellow"), size = 0.4) +
  geom_line(aes(y = brate_3539, color = "cyan"), size = 0.4) +
  geom_line(aes(y = brate_4044, color = "green"), size = 0.4) +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_1519, label = "Age 15-19"), x = 1980, y = 60, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_2024, label = "Age 20-24"), x = 1980, y = 100, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_2529, label = "Age 25-29"), x = 1980, y = 125 , hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_3034, label = "Age 30-34"), x = 1980, y = 80, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_3539, label = "Age 35-39"), x = 1980, y = 25, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_4044, label = "Age 40-44"), x = 1980, y = 5, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_vline(xintercept = 2007, linetype = "dotted") +
  annotate("text", x = 2011, y = 122, label = "2007", vjust = -0.5, hjust = 0.5, color = "black", family = "Times") +
  geom_hline(yintercept = seq(0, 140, by = 20), color = "lightgrey") +
  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 140, by = 20), labels = seq(0, 140, by = 20)) +
  scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 1), labels = function(x) ifelse(x %% 5 == 0, x, "")) +
  labs(x = NULL, 
     y = 'Births per 1,000 women in relevant population subgroup',
     title = "A: Five-year age group",
     caption = NULL,
     family = "Times") +
theme(axis.title.y = element_text(size = 4, family = "Times")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = -45, family = "Times", size = 8),
        plot.title = element_text(size = 10, family = "Times"),
        legend.position = "none")

@@

For future research, exploring additional aspects of society such as political demographics, immigration law, and enforcement, based on the trends observed in @fig-polscatter-graph, could provide valuable insights. Understanding the potential impact of immigration laws and campaign promises during presidential mandates on birth rates could enhance our understanding of the factors influencing demographic trends. Further exploration of these topics could contribute to a deeper understanding of the broader societal context surrounding birth rate trends in the United States

>>>>

---
title: "Potential Factors of Decreasing Birth Rates in the US since the Great Recession and Beyond"
subtitle: "Reproduction of 'The Puzzle of Falling US Birth Rates since the Great Recession'(Kearney, Levine & Pardue, 2022)"
author: 
  - Chay Park
thanks: "Code and data are available at: https://github.com/Chay-HyunminPark/Social-Science-Study. A replication of various aspects in this paper are available at: https://doi.org/10.48152/ssrp-55pn-6h10"
date: today
date-format: long
abstract: "This paper replicates Kearney, Levine, and Pardue's (2022) study on the notable decrease in US birth rates from 2007 to 2020. Analyzing data spanning from 1980 to 2020, we examine variations in birth rates among different age groups and disparities based on the race and ethnicity of mothers. Our analysis uncovers additional factors contributing to this trend beyond the initial attribution to the Great Recession. Specifically, socio-economic factors such as contraceptive use, cultural norms, and financial stability play key roles in the observed decline in birth rates."
format: pdf
number-sections: true
bibliography: references.bib
prefer-html: true
---

@@

According to the Centers for Disease Control and Prevention (CDC), the birth rate is defined as "the number of live births per 1,000 population." This rate is calculated by dividing the number of live births in a population in a year by the midyear resident population. Specifically, for census years, rates are calculated based on the unrounded census counts of the resident population as of April 1st.

From an economic perspective, birth rates are a vital determinant of a country's growth. Recent years have seen widespread concern over declining birth rates globally, prompting nations to explore strategies to address this issue. For example, in 2024, Germany's birth rate is reported as 9.3, lower than Canada's 10.0 and the US's 12.0 (Macrotrends 2024). Germany is taking action by actively recruiting skilled workers from abroad (Grunau 2023) and reforming its citizenship law to allow dual citizenship (Knight 2024) as measures to combat declining birth rates and overall population decline.

Furthermore, birth rates serve as an indicator of replacement-level fertility, essential for maintaining population stability in developed nations. Replacement-level fertility is the level of fertility at which each generation is exactly replaced by the next generation without any population growth or decline. It's typically estimated to be around 2.1 children per woman in developed countries, accounting for factors like infant mortality and individuals who don't reproduce. When birth rates fall below this threshold, it can lead to population decline, aging populations, and potential economic challenges.

The estimand, or the target of estimation, in this context is the birth rate itself – specifically, the average number of live births per 1,000 population over a specified period. Understanding and accurately estimating this estimand is crucial for assessing population dynamics and informing policy decisions aimed at addressing declining birth rates.

@@

This paper utilizes a replication package of the original study and employs a cleaned data CSV file to reproduce the related figures and tables. While the original code in the replication package was written in Stata, this paper utilizes R (R Core Team 2020) for analysis, leveraging packages such as ggplot2 (Wickham et al. 2016) for creating graphical representations. Additionally, the gridExtra package was used to enhance the layout and presentation of the figures (Murrell 2021).

The paper proceeds by first presenting the trend in US birth rates over the period 1980 to 2020. It then delves into variations in birth rates by age group and race/ethnicity of mothers, correlating these trends with education, socio-economic factors, and financial stability. The replication of Kearney, Levine, and Pardue's study focuses on answering the following research questions:

@@

In this paper, data validation is conducted using an R script file named '03-test_data' to ensure the accuracy of the dataset obtained directly from the original package. The paper utilizes both 'fig_1' and 'fig_2a_2b' CSV files.

In the dataset, 'brate_all' represents the birth rates of all women aged 15-44 in the US. 'Brate' is an abbreviation for 'birth rates'. For example, 'brate_1519' corresponds to the birth rate data for women aged 15 to 19 years old, irrespective of race and ethnicity. Similarly, 'brate_whitenh' represents the birth rates of white, non-Hispanic women, regardless of age.

Regarding race and ethnicity, "white" refers to the non-Hispanic Caucasian population group, "black" refers to the non-Hispanic African American group, and "Hispanic" refers to individuals with Hispanic ethnicity, primarily from South America, for brevity.

@@

# Read the data from CSV file
data <- read.csv(file = here::here("inputs/data/fig_1.csv"))

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-polscatter-graph
#| fig-cap: "Trend in US Birth Rates During the Presidential Mandate"

@@

# Read the data from CSV file
data <- read.csv(file = here::here("inputs/data/figs_2a_2b.csv"))

# Plotting the first line graph (Graph A)
graph_A <- ggplot(data, aes(x = year)) +
  geom_line(aes(y = brate_1519, color = "blue"), size = 0.4) +
  geom_line(aes(y = brate_2024, color = "orange"), size = 0.4) +
  geom_line(aes(y = brate_2529, color = "grey"), size = 0.4) +
  geom_line(aes(y = brate_3034, color = "yellow"), size = 0.4) +
  geom_line(aes(y = brate_3539, color = "cyan"), size = 0.4) +
  geom_line(aes(y = brate_4044, color = "green"), size = 0.4) +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_1519, label = "Age 15-19"), x = 1980, y = 60, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_2024, label = "Age 20-24"), x = 1980, y = 100, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_2529, label = "Age 25-29"), x = 1980, y = 125 , hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_3034, label = "Age 30-34"), x = 1980, y = 80, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_3539, label = "Age 35-39"), x = 1980, y = 25, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_text(data = data[data$year == 2020,], aes(x = year, y = brate_4044, label = "Age 40-44"), x = 1980, y = 5, hjust = -0.1, vjust = -0.5, size = 3, family = "Times") +
  geom_vline(xintercept = 2007, linetype = "dotted") +
  annotate("text", x = 2011, y = 122, label = "2007", vjust = -0.5, hjust = 0.5, color = "black", family = "Times") +
  geom_hline(yintercept = seq(0, 140, by = 20), color = "lightgrey") +
  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 140, by = 20), labels = seq(0, 140, by = 20)) +
  scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 1), labels = function(x) ifelse(x %% 5 == 0, x, "")) +
  labs(x = NULL, 
     y = 'Births rates in relevant population subgroup',
     title = "A: Five-year age group",
     caption = NULL,
     family = "Times") +
theme(axis.title.y = element_text(size = 4, family = "Times")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = -45, family = "Times", size = 8),
        plot.title = element_text(size = 10, family = "Times"),
        legend.position = "none")

@@

For future research, exploring additional aspects of society such as political demographics, immigration law, and enforcement, based on the trends observed in @fig-polscatter-graph, could provide valuable insights. Understanding the potential impact of immigration laws and campaign promises during presidential mandates on birth rates could enhance our understanding of the factors influencing demographic trends. Further exploration of these topics could contribute to a deeper understanding of the broader societal context surrounding birth rate trends in the United States.
